Replacing standard galaxy profiles with \\ mixtures of Gaussians

### 2012-05-08 docs/mixture_models.tex
to-do list
structure like a paper, intro text, reference list started

\documentclass[12pt]{article}
\usepackage{amssymb,amsmath,mathrsfs}

\newcommand{\documentname}{\textsl{Note}}
\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\tractor}{\project{Tractor}}

@@

\section*{Replacing standard galaxy profiles with \\ mixtures of Gaussians}

\noindent
David W. Hogg \\
\textsl{New York University} \\
\textsl{Max-Planck-Institut f\"ur Astronomie} \\[1ex]
Dustin Lang \\
\textsl{Princeton University Observatory}

@@

...Some other authors have worked with mixtures of Gaussians...Very
related work...thinking more generally...Here we want to solve a very
specific set of numerical problems...

...Mixtures of Gaussians sometimes used for PSF fitting...Used in
XD...Other places?...

...Rendering of Sersic profiles (of which exp and dev are types) is
non-trivial at the center...for the same reason convolution is a
bitch...

...If we wanted to approximate these models, what objective function
would we use?  What tolerances are okay?...

@@

...Define a sensible scalar objective...Give fitting procedure...give actual values for the parameters...

@@

>>>>

% to-do
% -----
% - zeroth draft
% - make useful figures
% - send to Jim Bosch and Kevin Bundy for comments

\documentclass[12pt,pdftex,preprint]{aastex}
\usepackage{amssymb,amsmath,mathrsfs}

\newcommand{\foreign}[1]{\textit{#1}}
\newcommand{\etal}{et\,al.}
\newcommand{\documentname}{\textsl{Note}}
\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\tractor}{\project{Tractor}}
\newcommand{\sdss}{\project{SDSS}}

@@

\title{Replacing standard galaxy profiles with \\ mixtures of Gaussians}
\author{David W. Hogg\altaffilmark{1,2,3} \&
        Dustin Lang\altaffilmark{4}}
\altaffiltext{1}{To whom correspondence should be addressed; \texttt{david.hogg@nyu.edu}}
\altaffiltext{2}{New York University}
\altaffiltext{3}{Max-Planck-Institut f\"ur Astronomie}
\altaffiltext{4}{Princeton University Observatory}

@@

Gaussians are remarkable distribution functions.  They have the
incredible properties that---in any number of dimensions---the
convolution (or correlation) of one multivariate Gaussian with another
is itself a multivariate Gaussian, and any product of multivariate
Gaussians is itself a multivariate Gaussian but with a different
normalization.  Furthermore, the means and variance tensors of the
Gaussians output by these operations are related simply to the means
and variance tensors of the inputs.  Add to these wonders the fact
that Gaussians form a complete basis for representing (smooth)
probability distribution functions and it becomes remarkable that we
don't do everything we do in terms of Gaussians.

To elaborate, a mixture of multivariate Gaussians---a linear
superposition---can be used to represent any reasonable distribution
in any number of dimensions to any reasonable precision.  Convolution
(or correlation) by any other distribution that has also been
represented by a mixture of multivariate Gaussians creates a new
mixture of Gaussians with simply adjusted amplitudes, means, and
variance tensors.  The ubiquity of convolution operations in astronomy
suggests the widespread adoption of mixture-of-Gaussian modeling.  We
have pioneered this in the area of distribution modeling in large
numbers of dimensions \citep{xd, xdqso, xdqsoz}, where convolution
occurs because the true or noise-free distribution is convolved with
the noise before being observed.  Here we are going to capitalize on
the convolution properties of mixtures of Gaussians in modeling galaxy
morphologies in imaging data, where the true or
high-angular-resolution intensity field is convolved with the
point-spread function (PSF) before being observed.  This is
particularly useful in data sets in which the PSF itself has also been
modeled as a mixture of Gaussians, which is not uncommon.  For
example, the \project{Sloan Digital Sky Survey} (\sdss; [HOGG cite
  York]) imaging pipelines [HOGG cite Lupton] makes such models for
every imaging field.

We are not the first in this space....modeling galaxies with mixtures
of Gaussians has been done very successfully before [HOGG cite].
However, the idea behind that work was to use the mixture of Gaussians
to provide a very free form for the isophotes of large, nearby, highly
resolved galaxies.  Here our goals are very limited: We want to
improve the performance of standard galaxy image model fitting by
expressing the standard galaxy models---the exponential and
de~Vaucouleurs profiles---as rigid mixtures of Gaussians.

The performance advantages we will obtain from using mixtures of
Gaussians are not simply that convolution itself is trivial.  The
standard unconvolved galaxy models, especially the de~Vaucouleurs and
S\'ersic models, are very ill-behaved near the galaxy center.
Rendering these profiles precisely near the center can be very
challenging numerically.  In addition, most of the rendering time is
spent at very small radii, where PSF convolution is going to erase all
structure anyway.  That is, a mixture-of-Gaussians description of the
de~Vaucouleurs profile saves time both in rendering \emph{and} in PSF
convolution; it produces profiles that are approximate but very high
performance when all real uses of the profiles are PSF-convolved, as
they usually are.

@@

...If we wanted to approximate these models, what objective function
would we use?  What tolerances are okay?...Define a sensible scalar
objective...Give fitting procedure...give actual values for the
parameters...

@@

\acknowledgements
It is a pleasure to thank...

\begin{thebibliography}{70}
\bibitem[Bovy \etal(2011a)]{xd}
Bovy,~J., Hogg,~D.~W., \& Roweis,~S., 2011, Ann. Appl. Stat., 5, 1657
\bibitem[Bovy \etal(2011b)]{xdqso}
Bovy,~J., \etal, 2011, \apj, 729, 141
\bibitem[Bovy \etal(2012)]{xdqsoz}
Bovy,~J. \etal, 2012, \aj, 749, 41
\end{thebibliography}
